---
title: "Queensland Address Density Analysis"
author: "Ashlee Poeppmann"
date: "2026-08-03"
output: html_document
---

## Objective

Which localities in Queensland have the highest density of addresses relative to land area?

#will need to change to just SEQ (cut area in QSpatial) and test upload to Github.

## Data Sources

Need to add in where got from and metatdata and usage etc

- Property Address Queensland (text data package \| GDA1994 EPSG:4283)

- Locality Boundaries — Queensland (geodatabase \| GDA2020 EPSG:7844)

- Some of the analysis I used

## Definitions / Assumptions

Localities = Suburbs

## Methodology

### 1 Loading and Preparing the Data

```{r setup, include=FALSE}
#Libraries
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(leaflet)
library(ggrepel)
library(stringr)
```

#### 1.1 Queensland Locality Boundaries

This data is supplied as a gdb, and already has it's datum set.

```{r load locality boundaries}
localities_sf <- st_read("data/LocalityBoundaries.gdb", layer = "Locality_Boundaries")
```

Queensland has a small number of locality names that repeat across different local government areas, Example: West End (Brisbane City) and West End (Townsville City). We'll need to join these to avoid duplication.

```{r locality LGA}
localities_sf <- localities_sf %>%
  mutate(locality_base = str_remove(locality, " \\(.*\\)$"))

dupe_names <- localities_sf %>%
  st_drop_geometry() %>%
  count(locality_base) %>%
  filter(n > 1) %>%
  pull(locality_base)

localities_sf <- localities_sf %>%
  mutate(locality = if_else(
    locality_base %in% dupe_names,
    paste0(locality_base, " (", lga, ")"),
    locality_base
  ))
```

Checking how the data looks, and confirming the datum = GDA2020.

```{r check-localities}
# Quick check data loaded correctly
#total localities
nrow(localities_sf)
#Check Datum
st_crs(localities_sf)$input


#sanity check above (LGA) append worked
localities_sf %>%
  st_drop_geometry() %>%
  filter(str_detect(locality, "West End")) %>%
  select(locality, lga)

#Quick map view
localities_sf %>%
  ggplot() +
  geom_sf(fill = "lightblue", color = "darkblue", linewidth = 0.1) +
  theme_minimal() +
  labs(title = "Locality Boundaries")

```

#### 1.2 Property Address Queensland

The data is supplied in a txt document, which includes a column on what datum to use, so I have loaded it in as EPSG:4283 GDA94.

```{r load-addresses}
addresses_raw <- read_delim(
  "data/Property_address_Queensland.txt",
  delim = "|",
  col_types = cols(LATITUDE = col_double(), LONGITUDE = col_double(), .default = col_character())
)

addresses_sf <- st_as_sf(
  addresses_raw,
  coords = c("LONGITUDE", "LATITUDE"),
  crs = 4283,
  remove = FALSE
)
```

We also only want the Status = P not A, as P = Primary address. Alternative addresses in the data are 'second' addresses, usually reserved for corner blocks.

Locality boundary data is in GDA2020 while address data is in GDA94; all layers need to be transformed to a common CRS (GDA2020 EPSG:7844) prior to spatial analysis. GDA2020 is shiny, new and has a higher accuracy. Also future proofing for any future analysis with other layers that may be GDA2020.

```{r addresses formating}
#Only 'P' addresses
addresses_primary <- addresses_sf %>%
  filter(ADDRESS_STATUS == "P")

addresses_primary <- st_transform(addresses_primary, crs = 7844)

```

Checking how the data looks and confirming the datum was properly set.

```{r check-addresses}
#total addresses
nrow(addresses_primary)
#Check Datum
st_crs(addresses_primary)$input


#Quick map view of a sample amount
addresses_primary %>%
  slice_sample(n = 5000) %>%
  ggplot() +
  geom_sf(size = 0.3, alpha = 0.3) +
  theme_minimal() +
  labs(title = "Sample of Primary Address points (n=5000)")
```

### 2. Spatial Join

Taking each point in the address layer and finds which polygon (locality) it falls inside, then attaches that polygon's attributes to the point.

```{r spatial join}
#Spatial join 'within' - point must fall 'within' the polygon
addresses_joined <- st_join(
  addresses_primary,
  localities_sf,
  join = st_within
)
```

Sanity Checking the join

```{r join validation}
#Number of rows from addresses
nrow(addresses_primary)
#nrow(addresses_joined)
nrow(addresses_joined)
#These two numbers should match if every point match
```

Result = 2739452. Both row count matches, meaning the join didn't create duplicates or drop any address points.

Sanity check - checking whether each address's self-reported locality (from the original text file) agrees with the spatially-derived locality (form the join above)

```{r data quality checking}
addresses_joined <- addresses_joined %>%
  mutate(
    LOCALITY_stripped = str_remove(LOCALITY, " \\(.*\\)$"),
    locality_stripped = str_remove(locality, " \\(.*\\)$"),
    locality_match = toupper(LOCALITY_stripped) == toupper(locality_stripped) #toupper change case
  )

table(addresses_joined$locality_match, useNA = "ifany")
mean(addresses_joined$locality_match, na.rm = TRUE) * 100
```

**4 points** = N/A, meaning these points didn't fall inside any locality polygon at all (likely just outside the Queensland border or off the coast).

**2734925 points (99.83%**) = the address record's self-reported locality matches the locality polygon the point geographically falls inside

**4523 points (\~0.17%)** = a mismatch between the two sources

**Result** = \~99.38% agreement rate between two fields

What boundaries causes the \~0.62% mismatch?

```{r mismatch calc}

mismatches <- addresses_joined %>%
  filter(locality_match == FALSE) %>%
  st_drop_geometry() %>%
  count(LOCALITY, locality, sort = TRUE) %>%
  head(20)

mismatches

```

Most of these are data quality errors that are out of scope to clean now. One example is BELLS CREEK vs Aura Central. [Aura Central](https://haveyoursay.nrmmrrd.qld.gov.au/bells-creek-and-aura-central-place-name-proposal) is a newly created locality, as of 25th July. Addresses have not been updated yet. I'm happy with 99.38% result.

How many addresses fall in each locality?

```{r no. addresses in locality}
address_counts <- addresses_joined %>%
  st_drop_geometry() %>%
  filter(!is.na(locality)) %>%
  count(locality, name = "address_count")

head(address_counts, 10)
```

### 3. Area/density calculation

Density calculation — joining address counts to each locality's land area, then computing addresses per km2

```{r density}
locality_density <- address_counts %>%
  left_join(
    localities_sf %>% st_drop_geometry() %>% select(locality, ca_area_sqkm),
    by = "locality"
  ) %>%
  mutate(density_per_sqkm = address_count / ca_area_sqkm)

locality_density %>% arrange(desc(density_per_sqkm)) %>% head(10)
```

Top 10 localities ranked by address density per km2. Note: I've added a filter(address_count \>= 100) to exclude any locality with fewer than 100 addresses, so a tiny locality with a handful of addresses on a very small locality can't produce an artificially extreme density value.

```{r - top 10 localities ranked by address density per km2}
locality_density %>%
  filter(address_count >= 100) %>%
  arrange(desc(density_per_sqkm)) %>%
  select(locality, address_count, ca_area_sqkm, density_per_sqkm) %>%
  head(10)
```

A note for future configurations and fine tuning: joining the Local Government Areas to the above table, might help with a quick analysis of where the localities are in QLD.

### 4. Visualize

####4.1 Charts

Bar charts

```{r bar chart}
locality_density %>%
  arrange(desc(density_per_sqkm)) %>%
  head(10) %>%
  ggplot(aes(x = reorder(locality, density_per_sqkm), y = density_per_sqkm)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(density_per_sqkm, 0)), hjust = -0.1, size = 3) +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Top 10 Localities by Address Density",
    x = NULL,
    y = "Addresses per km2"
  )
```

A scatterplot - log scale is used sic=nce most of the data is bunched at the low end.

```{r scatterplot}

area_count_outliers <- locality_density %>%
  filter(address_count > 15000 | ca_area_sqkm > 20000)

area_count_outliers %>% select(locality, address_count, ca_area_sqkm)

locality_density %>%
  ggplot(aes(x = ca_area_sqkm, y = address_count)) +
  geom_point(alpha = 0.4, color = "steelblue") +
  geom_point(data = area_count_outliers, color = "red", size = 2) +
  geom_text_repel(data = area_count_outliers, aes(label = locality), size = 3) +
  scale_x_log10() +
  theme_minimal() +
  labs(
    title = "Address Count by Locality Land Area",
    x = "Land Area (km2, log scale)", y = "Address Count"
  )
```

####4.2 Maps Preparing data for map — joining calculated density statistics back onto the spatial polygons.Leaflet, like many web-based mapping tools and services — including Esri Feature Services, which I've worked with directly — requires coordinates in WGS84 (EPSG:4326) to render data correctly.

```{r leaflet set up}
localities_map_data <- localities_sf %>%
  left_join(
    locality_density %>% st_drop_geometry(),
    by = "locality"
  ) %>%
  filter(address_count >= 100) %>%
  st_transform(crs = 4326)  # WGS84, required for leaflet
```

Map building, centered on Brisbane as a whole state view doesn't show the high density places.

```{r leaflet map}
pal <- colorNumeric(palette = "YlOrRd", domain = localities_map_data$density_per_sqkm, na.color = "transparent")

leaflet(localities_map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = 153.05, lat = -27.60, zoom = 9) %>%  # centered on South East Queensland
  addPolygons(
    fillColor = ~pal(density_per_sqkm),
    fillOpacity = 0.7,
    color = "white",
    weight = 0.5,
    label = ~paste0(locality, ": ", round(density_per_sqkm, 1), " addr/km²") %>% lapply(htmltools::HTML)
  ) %>%
  addLegend(pal = pal, values = ~density_per_sqkm, title = "Addresses per km²", position = "bottomright")
```

## Findings

## Reflections
