---
title: "map_address_density_analys"
output: html_document
date: "2026-08-03"
output:
  html_document:
    toc: true
---


Loading and Preparing the Data

```{r setup, include=FALSE}
#Libraries
library(sf) #reading data
library(dplyr)
library(readr)
library(ggplot2)
library(leaflet)
library(ggrepel)
library(stringr)
```

#### Queensland Locality Boundaries

```{r}
# Reading directly from an ArcGIS Feature Service REST endpoint
localities_web <- read_sf("https://spatial-gis.information.qld.gov.au/arcgis/rest/services/Boundaries/AdministrativeBoundaries/MapServer/2/query?where=lga+IN+%28%27Brisbane+City%27%2C+%27Gold+Coast+City%27%2C+%27Somerset+Regional%27%2C+%27Sunshine+Coast+Regional%27%2C+%27Moreton+Bay+City%27%2C+%27Redland+City%27%2C+%27Logan+City%27%2C+%27Noosa+Shire%27%2C+%27Scenic+Rim+Regional%27%2C+%27Ipswich+City%27%2C+%27Lockyer+Valley+Regional%27%2C+%27Toowoomba+Regional%27%29&outFields=*&returnGeometry=true&f=geojson")

```

Queensland has a number of locality names that repeat across different local government areas, Example: Tamborine (Logan City) and Tamborine (Scenic Rim Regional). We'll need to join these to avoid duplication in South East Queensland.

```{r locality LGA}
localities_web <- localities_web %>%
  mutate(locality_base = str_remove(locality, " \\(.*\\)$"))

dupe_names <- localities_web %>%
  st_drop_geometry() %>%
  count(locality_base) %>%
  filter(n > 1) %>%
  pull(locality_base)

localities_web <- localities_web %>%
  mutate(locality = if_else(
    locality_base %in% dupe_names,
    paste0(locality_base, " (", lga, ")"),
    locality_base
  ))

#Force correct crs - in EPSG:4326 but we want 3857 for spatial data analysis
localities_web <- st_transform(localities_web, crs = 3857)

```


#### Property Address Queensland

The text data download includes a column on what datum to use, so I have loaded it in as EPSG:4283 GDA94. Plus I will convert the LOCAL_AUTHORITY field to sentence case sine they are all in capitals

```{r load-addresses}
#Read in text file and state delimeter.
addresses_raw <- read_delim(
  "data/Property_address_Queensland.txt",
  delim = "|",
  col_types = cols(LATITUDE = col_double(), LONGITUDE = col_double(), .default = col_character())
)

#add lat and longs from table
addresses_sf <- st_as_sf(
  addresses_raw,
  coords = c("LONGITUDE", "LATITUDE"),
  crs = 4283,
  remove = FALSE
)

#change casing of LOCAL_AUTHORITY attributes
addresses_sf <- addresses_sf %>%
  mutate(LOCAL_AUTHORITY = str_to_title(LOCAL_AUTHORITY))
```


```{r addresses formating}
#list of SEQ LGAs
seq_lgas <- c(
  "Brisbane City", "Gold Coast City", "Somerset Regional",
  "Sunshine Coast Regional", "Moreton Bay City", "Redland City",
  "Logan City", "Noosa Shire", "Scenic Rim Regional",
  "Ipswich City", "Lockyer Valley Regional", "Toowoomba Regional"
)

addresses_seq <- addresses_sf %>%
  filter(LOCAL_AUTHORITY %in% seq_lgas)

#Only 'P' addresses
addresses_primary <- addresses_seq %>%
  filter(ADDRESS_STATUS == "P")

#update crs
addresses_primary <- st_transform(addresses_primary, crs = 3857)

```


```{r}

pal <- colorNumeric(palette = "YlOrRd", domain = localities_map_data$density_per_sqkm, na.color = "transparent")

leaflet(localities_map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  setView(lng = 153.05, lat = -27.60, zoom = 9) %>%  # centered on South East Queensland
  addPolygons(
    fillColor = ~pal(density_per_sqkm),
    fillOpacity = 0.7,
    color = "white",
    weight = 0.5,
    label = ~paste0(locality, ": ", round(density_per_sqkm, 1), " addr/km²") %>% lapply(htmltools::HTML)
  ) %>%
  addLegend(pal = pal, values = ~density_per_sqkm, title = "Addresses per km²", position = "bottomright")
```

